Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2022-03-29 2022/eamt/eamt-latex/eamt22.tex
full review

\newcommand{\vlambda}{\ensuremath{\boldsymbol\lambda}\xspace} % parameters vector for a distribution
\newcommand{\vtheta}{\ensuremath{\boldsymbol\theta}\xspace} % parameters vector for a distribution
\newcommand{\vpsi}{\ensuremath{\boldsymbol\psi}\xspace} % parameters vector for a distribution
\newcommand{\indic}[1]{\ensuremath{\mathbb{I}(#1)}}
% \newcommand{\SB}[1]{\textcolor{green}{#1}}
% \newcommand{\SW}[1]{\textcolor{red}{#1}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}
% limits underneath
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\renewcommand\textfraction{.1}
\renewcommand\floatpagefraction{.95}
\newcommand{\sbcl}[2]{{\scriptsize #1 \hfill $|$ \hfill  #2}}
\usepackage{multirow}
\usepackage{adjustbox}
\renewcommand{\floatpagefraction}{.9}

@@

Several recent proposals
% \cite{Vanderwees17dynamic,Graves17automated,platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Wang20balancing,Wang20learning-multi}
explore ways to instead consider \emph{dynamic} data selection and sampling strategies: \newcite{Vanderwees17dynamic,Zhang19curriculum} construct a static curriculum, while \newcite{Graves17automated,Platanios19competence,Kumar19reinforcement,Wang20learning-multi,Wang20balancing} build curricula that automatically adapt to the training data. In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing}, initially applied to multilingual NMT, that simultaneously accounts for the domain adaptation and the multidomain adaptation problems.
	\item We show that MDAC achieves performance that compare to fine-tuning strategies for DA (\textsection~\ref{ssec:da}) and outperform some static data sampling strategies for multidomain settings (\ref{ssec:mda}).
	\item We show that our variant MDAC mitigates some failures of DDS in multidomain training.
	\item We illustrate the generality of differentiable data selection frameworks (both MDAC and DDS) on less common situations such as DA using unsupervised clustering (\textsection~\ref{ssec:clda}); DA using out-of-domain training data and small in-domain validation data (\textsection~\ref{ssec:uda}); and two-domain adaptation where the test distribution only mixes two of the training domain (\textsection~\ref{ssec:bida}).
\end{itemize}

@@

\section{Experimental settings} \label{sec:exp}
\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French in 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \cite{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Additional experiments use the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu} \cite{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

\begin{table*}[htbp]
  \centering
  \begin{tabular}{|l|ccccccc|} %*{4}{|r|}}
    \cline{2-8} 
    %\multicolumn{4}{|l|}{Vocab size - En: 30,165, Fr: 30,398}\\
    \multicolumn{1}{c|}{} & \multicolumn{1}{c}{\domain{med}} & \multicolumn{1}{c}{\domain{law}} & \multicolumn{1}{c}{\domain{bank}} & \multicolumn{1}{c}{\domain{it}} & \multicolumn{1}{c}{\domain{talk}} & \multicolumn{1}{c}{\domain{rel}} & \multicolumn{1}{c|}{\domain{news}} \\
    \hline 
    \# lines & 2609 (0.68) & 501 (0.13) & 190 (0.05) & 270 (0.07) & 160 (0.04) & 130 (0.03) & 260 (0) \\
    \# tokens &  133 / 154  &  17.1 / 19.6 &  6.3 / 7.3 &  3.6 / 4.6 &  3.6 / 4.0 &  3.2 / 3.4 & 7.8 / 9.2   \\
    \# types & 771 / 720 & 52.7 / 63.1 & 92.3 / 94.7 & 75.8 / 91.4 & 61.5 / 73.3 & 22.4 / 10.5 & - \\
    \# uniq & 700 / 640 & 20.2 / 23.7 & 42.9 / 40.1 & 44.7 / 55.7 & 20.7 / 25.6 & 7.1 / 2.1 & - \\
    \hline
  \end{tabular}
  \caption{Corpora statistics: number of parallel lines ($\times 10^3$) and proportion in the training domain mixture (which does not contain \domain{news}), number English and French tokens ($\times 10^6$), number English and French types ($\times 10^3$), number of types that only appear in a given domain ($\times 10^3$). \domain{med} is the largest domain, containing almost 70\% of the sentences, while \domain{rel} is the smallest, with only 3\% of the data.
  }
\label{tab:Corpora}
\end{table*}

@@

\subsection{Experimental tasks}
We evaluate our method in the 5 following conditions.
%including: domain adaptation; multi-domain adaptation, bi-domain adaptation, unseen domain adaptation, and unsupervised domain adaptation.
In the \emph{supervised domain adaptation task}, given the data from 6 domains (\domain{med}, \domain{bank}, \domain{law}, \domain{it}, \domain{talk}, \domain{rel}), we aim to build expert NMT models for each domain. To challenge the flexibility of the method, we also consider a \emph{two-domain adaptation task}, where given the same 6~domains, we focus on adapting to a mixture of 2~domains.

In the \emph{multidomain adaptation task}, we use the same 6 domains to build one single NMT model that should perform optimally, assuming a uniform distribution of domains during the test.

In a fourth experiment (\emph{unseen domain adaptation}), given training data in 6 domains and a small dev set of a new domain (\domain{News}), we build a model which performs well for the unseen domain.

Finally, in the \emph{unsupervised domain adaptation task}, we cluster all available training data into 30~clusters using the KNN algorithm in the same way as in \cite{Tars18multidomain}, then adapt these clusters to one of 6 domains using the corresponding in-domain dev set. We compare MDAC to DDS for each of our 6 test sets.

@@

\subsection{Domain Adaptation}\label{ssec:da}
In this setting, we aim to build an NMT model for one single domain: we accordingly set $\vlambda^t$ to a deterministic distribution $\vlambda_d$, where the target domain $d$ has probability~1.

@@

We now compare the best MDAC system (using $\vlambda_{t=0} = \vlambda_0$) to full fine-tuning. According to Table~\ref{tab:da}, fine-tuning is better for large domains such as \domain{med} and \domain{law}, while MDAC outperforms fine-tuning by approximately 1.2 BLEU for \domain{bank} and 1.0 BLEU for \domain{rel}. This indicates that for small domains, out-of-domain data helps improve the generalization and that MDAC is able to exploit both the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning. Results for DDS display similar trends but are always outperformed by MDAC. The behavior of CL, which does only well the large domain \domain{med} lag somewhat behind.

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{avg.} \\ \hline
    FT-Full($d$) &40.3&63.8&54.4&38.5&52.0&91.0&56.7\\ \hline
    \hline
    \system{CL($d$)} &40.2&60.2&53.7&36.5&51.1&91.1&55.5\\ \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} &39.6&60.1&55.0&38.5&52.5&92.0&56.3\\
    \system{MDAC($\vlambda_0, \vlambda_d$)} &39.6&62.5$^{**}$&55.6$^{*}$&38.5&52.4&92$^{***}$&56.8\\
     \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} &39.7&53.9&49.6&37.9&43.1&64.3&48.1\\
    \system{MDAC($\vlambda_1, \vlambda_d$)} &40.2&59.9&52.6&38.5&50.7&79.8&53.6\\ \hline
    \system{DDS($\vlambda_d, \vlambda_d$)} &39.9&63.9&54.5&35.4&51.2&91.8&56.1\\
    \system{MDAC($\vlambda_d, \vlambda_d$)}&40.6&63.9&54.5&35.6&51.3&92.3&56.4\\
    \hline
  \end{tabular}
  \caption{Domain adaptation experiments. We report BLEU scores of each method for 6 target domains and their average: each column corresponds to a distinct system. ($^{*}$) MDAC is significantly better than CL, fine-tuning and DDS with $p<0.05$. ($^{**}$) MDAC is significantly better than CL and DDS with $p<0.05$. ($^{***}$) MDAC is significantly better than CL, fine-tuning with $p<0.05$.}
  \label{tab:da}
\end{table*}

\subsection{Bi-domain adaptation}\label{ssec:bida}

In these control experiments, we showcase the flexibility of dynamic sampling and try to adapt to (arbitrary) pairs of target domains with equal weight and contrast our results with those of DDS. Results are in Table~\ref{tab:bi-da}. Here, MDAC significantly outperforms DDS in two settings (\domain{med+it} and \domain{law+bank}) while being surpassed in \domain{talk+rel}.
% Allthough we showing that MDAC is capable to adapt to a variable set of target domains.

@@

We now turn to a more realistic scenario and consider multi-domain NMT, which aims to train one single system with optimal performance averaged over 6 domains. This setting targets a uniform test distribution $\vlambda^t = \vlambda_0$. In this situation, CL \cite{Zhang19curriculum} does not apply. We therefore only contrast the performance of MDAC, DDS and several fixed training data distribution $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$, where $\vlambda_{\alpha}$ is defined according to equation~\eqref{mixture:trn}.

We again initialize MDAC and DDS with two distribution $\vlambda_0$ and $\vlambda_1$. According to Table~\ref{tab:multi-da}, MDAC achieves the best performance with initial (uniform) $\vlambda_0$. The same conclusion holds for DDS. For this configuration, MDAC outperforms static training distributions including $\big[ \vlambda_0, \vlambda_{0.75}, \vlambda_{1.0}\big]$ by a significant margin, and performs slightly better than $\big[ \vlambda_{0.25}, \vlambda_{0.5} \big]$. This indicates that MDAC allows us to skip the somewhat heuristic choice of the optimal training mixture.

A second observation is that MDAC again outperforms DDS by a wide margin (+1.5 BLEU on average); the only domain where DDS does better is \domain{med}. Figure~\ref{fig:sampling}, which displays the evolution of the training mixture in the course of training, helps to understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the sampling distribution quickly reaches a bi-domain mode in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the distribution computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger proportion of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This, however, only happens at the end of training, when some NMT models might already be close to their best performance for the small domains.

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*7{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} \\ \hline \hline
    \system{DDS($\vlambda_0, \vlambda_2$)}&39.5&-&-&-&50.1&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&39.1&-&-&-&51.8$^*$&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&60.8&53.3&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&61.9$^*$&54.5$^*$&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&-&-&37.9&-&91.3 \\ 
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&-&-&36.9&-&90.4 \\
    \hline
  \end{tabular}
  \caption{Adapting to two domains. For a given line, non empty columns correspond to the pair of target domains. ($^*$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:bi-da}
\end{table*}

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline \hline
    % Multidomain
    \system{Mixed-0} &38.6&59.3&53.7&37.3&51.0&90.4&55.1\\
    \system{Mixed-0.25}&38.9&59.6&53.3&37.6&50.5&90.6&55.1\\
    \system{Mixed-0.5}&39.0&60.2&52.5&38.5&51.9&90.3&55.4\\
    \system{Mixed-0.75}&39.4&59.9&51.9&38.8&50.0&87.6&54.6\\
    \system{Mixed-1}&40.3&59.5&49.8&36.4&49.0&80.0&52.5\\
    \hline \hline
    \system{DDS($\vlambda_0, \vlambda_0$)} &40.1&56.9&50.7&37.4&46.8&92.0&54.0\\ 
    \system{MDAC($\vlambda_0, \vlambda_0$)}&38.5&60.3$^{**}$&54.4$^*$&37.3&51.3$^{**}$&91.4$^*$&55.5$^{**}$\\ 
    \hline \hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &40.6&55.5&48.0&36.2&46.9&60.1&47.9\\
    \system{MDAC($\vlambda_1, \vlambda_0$)}&40.2&59.3$^{**}$&51.0$^{**}$&36.9$^{**}$&48.6$^{**}$&80.7$^{**}$&52.8$^{**}$\\
    \hline
  \end{tabular}
  \caption{Multi domain adaptation. For a given line, all the columns \emph{correspond to the same multi-domain system}. ($^*$) MDAC is significantly better than \system{Mixed-$\alpha$} with $p<0.05$. ($^{**}$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:multi-da}
\end{table*}

\begin{figure*}[htbp]
\begin{subfigure}{.5\textwidth}
  \centering
  % include first image
  \includegraphics[width=0.97\textwidth]{DDS.png}  
  \caption{DDS}
  \label{fig:DDS}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  % include second image
  \includegraphics[width=0.97\textwidth]{MDAC.png}  
  \caption{MDAC}
  \label{fig:MDAC}
\end{subfigure}
\caption{Evolution of the sampling distribution during training.}
\label{fig:sampling}
\end{figure*}

>>>>

\newcommand{\vlambda}{\ensuremath{\boldsymbol\lambda}\xspace} % parameters vector for a distribution
\newcommand{\vtheta}{\ensuremath{\boldsymbol\theta}\xspace} % parameters vector for a distribution
\newcommand{\vpsi}{\ensuremath{\boldsymbol\psi}\xspace} % parameters vector for a distribution
\newcommand{\indic}[1]{\ensuremath{\mathbb{I}(#1)}}
% \newcommand{\SB}[1]{\textcolor{green}{#1}}
% \newcommand{\SW}[1]{\textcolor{red}{#1}}
\newcommand{\SB}[1]{\textbf{#1}}
\newcommand{\SW}[1]{\underline{#1}}
% limits underneath
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\renewcommand\textfraction{.1}
\renewcommand\floatpagefraction{.95}
\newcommand{\sbcl}[2]{{\scriptsize #1 \hfill $|$ \hfill  #2}}
\usepackage{multirow}
\usepackage{adjustbox}

@@

Several recent proposals
% \cite{Vanderwees17dynamic,Graves17automated,platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Wang20balancing,Wang20learning-multi}
explore ways to instead consider \emph{dynamic} data selection and sampling strategies: \newcite{Vanderwees17dynamic,Zhang19curriculum} construct a static curriculum, while \newcite{Wang20learning-multi,Wang20balancing} build curricula that automatically adapt to the training data. In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing}, initially applied to multilingual NMT, that simultaneously accounts for the domain adaptation and the multidomain adaptation problems.
	\item We show that MDAC achieves performance that compare to fine-tuning strategies for DA (\textsection~\ref{ssec:da}) and outperform some static data sampling strategies for multidomain settings (\ref{ssec:mda}).
	\item We show that our variant MDAC mitigates some failures of DDS in multidomain training.
	\item We illustrate the generality of differentiable data selection frameworks (both MDAC and DDS) on less common situations such as DA using unsupervised clustering (\textsection~\ref{ssec:clda}); DA using out-of-domain training data and small in-domain validation data (\textsection~\ref{ssec:uda}); and two-domain adaptation where the test distribution only mixes two of the training domain (\textsection~\ref{ssec:bida}).
\end{itemize}

@@

\section{Experimental settings} \label{sec:exp}
\subsection{Data and metrics \label{ssec:corpora}}
\begin{table*}[htbp]
  \vspace{-\baselineskip}
  \centering \footnotesize
  \begin{tabular}{|l|ccccccc|} %*{4}{|r|}}
    \cline{2-8} 
    %\multicolumn{4}{|l|}{Vocab size - En: 30,165, Fr: 30,398}\\
    \multicolumn{1}{c|}{} & \multicolumn{1}{c}{\domain{med}} & \multicolumn{1}{c}{\domain{law}} & \multicolumn{1}{c}{\domain{bank}} & \multicolumn{1}{c}{\domain{it}} & \multicolumn{1}{c}{\domain{talk}} & \multicolumn{1}{c}{\domain{rel}} & \multicolumn{1}{c|}{\domain{news}} \\
    \hline 
    \# lines & 2609 (0.68) & 501 (0.13) & 190 (0.05) & 270 (0.07) & 160 (0.04) & 130 (0.03) & 260 (0) \\
    \# tokens &  133 / 154  &  17.1 / 19.6 &  6.3 / 7.3 &  3.6 / 4.6 &  3.6 / 4.0 &  3.2 / 3.4 & 7.8 / 9.2   \\
    \# types & 771 / 720 & 52.7 / 63.1 & 92.3 / 94.7 & 75.8 / 91.4 & 61.5 / 73.3 & 22.4 / 10.5 & - \\
    \# uniq & 700 / 640 & 20.2 / 23.7 & 42.9 / 40.1 & 44.7 / 55.7 & 20.7 / 25.6 & 7.1 / 2.1 & - \\
    \hline
  \end{tabular}
  \caption{Corpora statistics: number of parallel lines ($\times 10^3$) and proportion in the training domain mixture (exluding \domain{news}), number English and French tokens ($\times 10^6$), types and uniq types ($\times 10^3$): the latter are types that only appear in a given domain.
    % \domain{med} is the largest domain, containing almost 70\% of the sentences, while \domain{rel} is the smallest, with only 3\% of the data.
  }
  \vspace{-\baselineskip}
\label{tab:Corpora}
\end{table*}

We experiment with translation from English into French in 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}) \cite{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \cite{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \cite{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \cite{Cettolo12wit}, and the Koran (\domain{rel}). Additional experiments use the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types, we use Byte-Pair Encoding \cite{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu} \cite{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

\subsection{Experimental tasks}
We evaluate our method in the 5 following conditions.
%including: domain adaptation; multi-domain adaptation, bi-domain adaptation, unseen domain adaptation, and unsupervised domain adaptation.
In the \emph{supervised domain adaptation task}, given the data from 6 domains (\domain{med}, \domain{bank}, \domain{law}, \domain{it}, \domain{talk}, \domain{rel}), we aim to build expert NMT models for each domain. To challenge the flexibility of the method, we also consider a \emph{two-domain adaptation task}, where given the same 6~domains, we focus on adapting to a mixture of 2~domains.
%
In the \emph{multidomain adaptation task}, we use the same 6 domains to build one single NMT model that should perform optimally, assuming a uniform distribution of domains during the test.
%
A fourth experiment (\emph{unseen domain adaptation}), adds to the training data for 6~domains a small dev set in a new domain (\domain{News}): our target is a model which performs well for the unseen domain.
%
Finally, in the \emph{unsupervised domain adaptation task}, we cluster all available training data into 30~clusters using the KNN algorithm as in \cite{Tars18multidomain}, then learn mixture weights these clusters to one of 6 domains using the corresponding dev set. We compare MDAC to DDS for each of our 6 test sets.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{avg.} \\ \hline
    FT-Full($d$) &40.3&63.8&54.4&38.5&52.0&91.0&56.7\\ \hline
    \hline
    \system{CL($d$)} &40.2&60.2&53.7&36.5&51.1&91.1&55.5\\ \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} &39.6&60.1&55.0&38.5&52.5&92.0&56.3\\
    \system{MDAC($\vlambda_0, \vlambda_d$)} &39.6&62.5$^{**}$&55.6$^{*}$&38.5&52.4&92$^{***}$&56.8\\
     \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} &39.7&53.9&49.6&37.9&43.1&64.3&48.1\\
    \system{MDAC($\vlambda_1, \vlambda_d$)} &40.2&59.9&52.6&38.5&50.7&79.8&53.6\\ \hline
    \system{DDS($\vlambda_d, \vlambda_d$)} &39.9&63.9&54.5&35.4&51.2&91.8&56.1\\
    \system{MDAC($\vlambda_d, \vlambda_d$)}&40.6&63.9&54.5&35.6&51.3&92.3&56.4\\
    \hline
  \end{tabular}
  \caption{Single domain adaptation. We report BLEU scores of each method for 6 target domains and their average: each column corresponds to a distinct system. ($^{*}$) MDAC is significantly better than CL, fine-tuning and DDS with $p<0.05$. ($^{**}$) MDAC is significantly better than CL and DDS with $p<0.05$. ($^{***}$) MDAC is significantly better than CL, fine-tuning with $p<0.05$.} \vspace{-\baselineskip}
  \label{tab:da}
\end{table*}
\subsection{Domain Adaptation}\label{ssec:da}
In this setting, we aim to build an NMT model for one single domain: we accordingly set $\vlambda^t$ to a deterministic distribution $\vlambda_d$, where the target domain $d$ has probability~1.

@@

We now compare the best MDAC system (using $\vlambda_{t=0} = \vlambda_0$) to full fine-tuning. According to Table~\ref{tab:da}, fine-tuning is better for large domains such as \domain{med} and \domain{law}, while MDAC outperforms fine-tuning by approximately 1.2 BLEU for \domain{bank} and 1.0 BLEU for \domain{rel}. This suggests that for small domains, out-of-domain data helps improve the generalization and that MDAC is able to exploit both the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning. Results for DDS display similar trends but are always outperformed by MDAC. Results for CL, which does only well the large domain \domain{med}, lag somewhat behind.

\subsection{Two-domain adaptation}\label{ssec:bida}

In these control experiments, we showcase the flexibility of dynamic sampling and adapt to (arbitrary) pairs of target domains with equal weight, contrasting MDAC with DDS in Table~\ref{tab:bi-da}. Here, MDAC significantly outperforms DDS in two settings (\domain{med+it} and \domain{law+bank}) out of three.
%while being surpassed in \domain{talk+rel}.
% Allthough we showing that MDAC is capable to adapt to a variable set of target domains.

@@

We now turn to a more realistic scenario and consider multidomain NMT, which aims to train one single system with optimal performance averaged over 6 domains and targets a uniform test distribution $\vlambda^t = \vlambda_0$. In this situation, CL \cite{Zhang19curriculum} does not apply: we only contrast the performance of MDAC, DDS and several fixed training data distribution $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$, where $\vlambda_{\alpha}$ is defined according to equation~\eqref{mixture:trn}.

We again initialize MDAC and DDS with two distribution $\vlambda_0$ and $\vlambda_1$. According to Table~\ref{tab:multi-da}, MDAC achieves the best performance with initial (uniform) $\vlambda_0$. The same conclusion holds for DDS. For this configuration, MDAC outperforms static training distributions including $\big[ \vlambda_0, \vlambda_{0.75}, \vlambda_{1.0}\big]$ by a significant margin, and performs slightly better than $\big[ \vlambda_{0.25}, \vlambda_{0.5} \big]$. Using MDAC thus dispenses with the empirical search of an optimal training mixture.

A second observation is that MDAC again outperforms DDS by a wide margin (+1.5 BLEU on average); the only domain where DDS does better is \domain{med}. Figure~\ref{fig:sampling}, which plots the evolution of the mixture weights during training, helps to understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the sampling distribution quickly reaches a bi-modal regime in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the distribution computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger part of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This only happens at the end of training, when some NMT models might already be close to their peak performance for the small domains.

\begin{table*}[htbp]
  \vspace{\baselineskip}
  \centering \small
  \begin{tabular}{|l|*7{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} \\ \hline \hline
    \system{DDS($\vlambda_0, \vlambda_2$)}&39.5&-&-&-&50.1&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&39.1&-&-&-&51.8$^*$&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&60.8&53.3&-&-&- \\
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&61.9$^*$&54.5$^*$&-&-&- \\
    \system{DDS($\vlambda_0, \vlambda_2$)}&-&-&-&37.9&-&91.3 \\ 
    \system{MDAC($\vlambda_0, \vlambda_2$)}&-&-&-&36.9&-&90.4 \\
    \hline
  \end{tabular}
  \caption{Adapting to two domains. For a given line, non empty columns correspond to the pair of target domains. ($^*$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:bi-da}
\end{table*}

\begin{table*}[htbp]
  \vspace{-2\baselineskip}
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline \hline
    % Multidomain
    \system{Mixed-0} &38.6&59.3&53.7&37.3&51.0&90.4&55.1\\
    \system{Mixed-0.25}&38.9&59.6&53.3&37.6&50.5&90.6&55.1\\
    \system{Mixed-0.5}&39.0&60.2&52.5&38.5&51.9&90.3&55.4\\
    \system{Mixed-0.75}&39.4&59.9&51.9&38.8&50.0&87.6&54.6\\
    \system{Mixed-1}&40.3&59.5&49.8&36.4&49.0&80.0&52.5\\
    \hline \hline
    \system{DDS($\vlambda_0, \vlambda_0$)} &40.1&56.9&50.7&37.4&46.8&92.0&54.0\\ 
    \system{MDAC($\vlambda_0, \vlambda_0$)}&38.5&60.3$^{**}$&54.4$^*$&37.3&51.3$^{**}$&91.4$^*$&55.5$^{**}$\\ 
    \hline \hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &40.6&55.5&48.0&36.2&46.9&60.1&47.9\\
    \system{MDAC($\vlambda_1, \vlambda_0$)}&40.2&59.3$^{**}$&51.0$^{**}$&36.9$^{**}$&48.6$^{**}$&80.7$^{**}$&52.8$^{**}$\\
    \hline
  \end{tabular}
  \caption{Multidomain adaptation. For a given line, all the columns \emph{correspond to the same multi-domain system}. ($^*$) MDAC is significantly better than \system{Mixed-$\alpha$} with $p<0.05$. ($^{**}$) MDAC is significantly better than DDS with $p<0.05$.}
  \label{tab:multi-da}
\end{table*}

\begin{figure*}[htbp]
  \vspace{-2\baselineskip}
  \begin{subfigure}{.5\textwidth}
  \centering
  % include first image
  \includegraphics[width=0.97\textwidth]{DDS.png}  
  \caption{DDS}
  \label{fig:DDS}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  % include second image
  \includegraphics[width=0.97\textwidth]{MDAC.png}  
  \caption{MDAC}
  \label{fig:MDAC}
\end{subfigure}
\caption{Evolution of the sampling distribution during training.}
\label{fig:sampling}
\end{figure*}
